---
title: "CoSparTA Pipeline Walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CoSparTA Pipeline Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette walks through the complete CoSparTA analysis pipeline on a 
synthetic clickstream dataset, from raw event logs to posterior inference 
and visualization. The pipeline has six stages: tensor preparation, model 
fitting, uncertainty quantification, covariate screening, post-processing, 
and visualization.

## Installation

Two dependencies are hosted on GitHub and must be installed first:

```{r install, eval = FALSE}
devtools::install_github("DongyueXie/ebpm")
devtools::install_github("DongyueXie/smashrgen")
devtools::install_github("nd-hal/CoSparTA")
```

```{r load}
library(CoSparTA)
```

## Stage 1: Build the Tensor

Raw data typically arrive as a long-format data frame with one row per 
observed event. `build_tensor()` converts this directly to the required 
$n \times p \times w$ integer count tensor in a single call, handling 
dimension labeling and ensuring all time bins are retained even if some 
are unobserved.

```{r build-tensor}
# Load the synthetic demo data
df <- readRDS(system.file("extdata", "demo_df.rds", package = "CoSparTA"))
cov_df <- read.csv(system.file("extdata", "demo_covariates.csv", 
                                package = "CoSparTA"))
channel_groups <- readRDS(system.file("extdata", "demo_channel_groups.rds", 
                                       package = "CoSparTA"))

# Reconstruct the universe of dimension labels
session_ids   <- sort(unique(df$session_id))
hour_labels   <- paste0("H", sprintf("%02d", 0:99))
channel_names <- c("Retail", "Social", "Review", "Deal", "Search")
website_names <- as.vector(
  outer(channel_names, paste0("_", sprintf("%02d", 1:10)), paste0)
)

# Build 1000 x 100 x 50 tensor
tensor_out <- build_tensor(
  data         = df,
  row          = "session_id",
  col          = "hour",
  slice        = "website",
  value        = "count",
  row_levels   = session_ids,
  col_levels   = hour_labels,
  slice_levels = website_names
)
X <- tensor_out$X  # 1000 x 100 x 50 integer tensor

# Covariate matrix: two continuous covariates
Xcov_mat <- as.matrix(cov_df[, c("cov1", "cov2")])
```

The resulting tensor has dimensions 1000 × 100 × 50 with sparsity approximately 0.97.

## Stage 2: Fit the Model

`CoSparTA()` is the main entry point. Supplying `Xcov` activates 
covariate-dependent priors on the observation mode, integrating contextual 
information directly into the decomposition rather than as a post-hoc 
adjustment. The rank $K$ reflects the assumed number of latent factors.

An optional CP-APR warm-start via `init_cpapr()` is recommended when Python 
is available, as it provides better initialization than random Gamma draws.

```{r fit-model}
# Optional CP-APR warm-start (requires Python with pyCP_APR)
# python3 -m venv cosparta_env
# source cosparta_env/bin/activate
# pip install pyCP_APR numpy
init_vals <- init_cpapr(X, K = 4, virtualenv = "cosparta_env")

# Covariate-aware fit
fit <- CoSparTA(
  X        = X,
  K        = 4,
  Xcov     = Xcov_mat,
  init     = init_vals,   # or "random_gamma" if Python unavailable
  maxiter  = 20,
  convergence_criteria = "factor_change",
  tol      = 1e-6,
  verbose  = TRUE
)

# Retrieve outputs via the unified getter
U1     <- get_loadings(fit, "U1")      # 1000 x 4 normalized loadings
U2     <- get_loadings(fit, "U2")      # 100  x 4 time factors
U3     <- get_loadings(fit, "U3")      # 50   x 4 channel weights
lambda <- get_loadings(fit, "weight")  # length-4 factor weights
gamma  <- get_loadings(fit, "gamma")   # list of 4 covariate coefficient vectors
```

When covariate effects are expected to differ across factors, rank-specific 
covariate sets can be supplied as a list. For instance, factor 1 receives 
both covariates, factor 2 is left covariate-free, while factors 3 and 4 
each receive a different covariate:

```{r rank-specific}
Xcov_cov1 <- Xcov_mat[, 1, drop = FALSE]
Xcov_cov2 <- Xcov_mat[, 2, drop = FALSE]

fit_rankspec <- CoSparTA(
  X    = X,
  K    = 4,
  Xcov = list(Xcov_mat, NULL, Xcov_cov1, Xcov_cov2),
  init = init_vals,
  convergence_criteria = "factor_change"
)
```

Missing entries are handled by `CoSparTA_missing()`, which restricts the 
likelihood to observed entries:

```{r missing}
mask     <- generate_missing_mask(X, missing_rate = 0.10)
fit_miss <- CoSparTA_missing(
  X        = X,
  K        = 4,
  Xcov     = Xcov_mat,
  obs_mask = mask$obs_mask,
  convergence_criteria = "factor_change"
)
```

## Stage 3: Uncertainty Quantification

Because the variational posterior retains the spike-and-slab structure of 
the prior, exact posterior credible intervals are available in closed form 
via `get_posterior_quantile()`.

```{r uq}
# Exact 95% posterior credible intervals for channel mode (U3)
ci_w <- get_posterior_quantile(fit, probs = c(0.025, 0.975), mode = "W")
# ci_w$q2.5, ci_w$q97.5  -- 50 x 4 lower and upper bounds

# Delta-method confidence intervals for covariate coefficients
ci_gamma <- get_gamma_ci(fit, level = 0.95)
# ci_gamma[[k]]$estimate, $se, $lower, $upper, $pvalue
print(ci_gamma[[1]])  # factor 1 results
```

## Stage 4: Covariate Screening

When the covariate set is large, `select_covariates()` provides an initial 
screening step. It fits a covariate-free model first, then regresses the 
observation-mode loadings on the covariates and flags those significant at 
$p < 0.05$ per factor. The selected covariates can then be passed to 
`CoSparTA()` via the list interface shown in Stage 2.

```{r screening}
# Step 1: fit covariate-free model
fit_unsup <- CoSparTA(
  X                    = X,
  K                    = 4,
  Xcov                 = NULL,
  init                 = init_vals,
  convergence_criteria = "factor_change"
)

# Step 2: screen covariates
sel <- select_covariates(
  K              = 4,
  covariate_data = as.data.frame(Xcov_mat),
  fit            = fit_unsup
)

# sel$selected: list of selected covariate indices per factor
print(sel$selected)
```

## Stage 5: Post-processing

`project_tensor()` scores new observations onto the estimated factor space 
without refitting the model. `reconstruct_tensor()` returns the denoised 
Poisson mean tensor $\hat{\mu}$.

```{r postprocessing}
# Project held-out sessions onto the factor space
X_new <- X[991:1000, , ]
F_new <- project_tensor(X_new, fit)  # 10 x 4 score matrix

# Reconstruct the denoised Poisson mean tensor
X_hat <- reconstruct_tensor(fit)     # 1000 x 100 x 50
```

## Stage 6: Visualization

Because non-negative CP decomposition is essentially unique 
(Kolda and Bader 2009), the extracted factor matrices carry a direct 
physical interpretation — each column represents a genuinely distinct 
latent behavioral pattern. Two dedicated plotting functions produce 
`ggplot2` figures for the time and channel factor matrices:

```{r visualization, fig.width = 7, fig.height = 4}
# Faceted line plot of K time factors
plot_time_factors(fit, time_labels = hour_labels)

# Faceted bar plot of K channel factors with category grouping
plot_channel_factors(fit,
                     channel_names  = website_names,
                     channel_groups = channel_groups)
```
